---
title: "Slides"
---


### Intro to Panel Data

```{r embed-xaringan1, echo=FALSE}
xaringanExtra::embed_xaringan(url = "slides/intro-cdc202108.html", ratio = "16:9")
```

<p><a class="btn btn-info btn-lg" target="_blank" href="slides/intro-cdc202108.html"><i class="fas fa-external-link-alt"></i> View all slides in new window</a> <a class="btn btn-info btn-lg" target="_blank" href="slides/intro-cdc202108.pdf" role="button"><i class="far fa-file-pdf"></i> Download PDF of all slides</a></p>



### Basics of DD

```{r embed-xaringan2, echo=FALSE}
xaringanExtra::embed_xaringan(url = "slides/did-cdc202108.html", ratio = "16:9")
```

<p><a class="btn btn-info btn-lg" target="_blank" href="slides/did-cdc202108.html"><i class="fas fa-external-link-alt"></i> View all slides in new window</a> <a class="btn btn-info btn-lg" target="_blank" href="slides/did-cdc202108.pdf" role="button"><i class="far fa-file-pdf"></i> Download PDF of all slides</a></p>


### TWFE and Event Studies

```{r embed-xaringan3, echo=FALSE}
xaringanExtra::embed_xaringan(url = "slides/twfe-cdc202108.html", ratio = "16:9")
```

<p><a class="btn btn-info btn-lg" target="_blank" href="slides/twfe-cdc202108.html"><i class="fas fa-external-link-alt"></i> View all slides in new window</a> <a class="btn btn-info btn-lg" target="_blank" href="slides/twfe-cdc202108.pdf" role="button"><i class="far fa-file-pdf"></i> Download PDF of all slides</a></p>


### Recent DD Advancements

```{r embed-xaringan4, echo=FALSE}
xaringanExtra::embed_xaringan(url = "slides/newdd-cdc202108.html", ratio = "16:9")
```

<p><a class="btn btn-info btn-lg" target="_blank" href="slides/newdd-cdc202108.html"><i class="fas fa-external-link-alt"></i> View all slides in new window</a> <a class="btn btn-info btn-lg" target="_blank" href="slides/newdd-cdc202108.pdf" role="button"><i class="far fa-file-pdf"></i> Download PDF of all slides</a></p>



### New Panel Methods

```{r embed-xaringan5, echo=FALSE}
xaringanExtra::embed_xaringan(url = "slides/newpanel-cdc202108.html", ratio = "16:9")
```

<p><a class="btn btn-info btn-lg" target="_blank" href="slides/newpanel-cdc202108.html"><i class="fas fa-external-link-alt"></i> View all slides in new window</a> <a class="btn btn-info btn-lg" target="_blank" href="slides/newpanel-cdc202108.pdf" role="button"><i class="far fa-file-pdf"></i> Download PDF of all slides</a></p>

